---
title: "3. Data Structure - Part 1"
description: |
   In this lesson we will review different data structures in R, including vectors, matrices, and arrays.
# author:
#   - name: Jenny Sloane
date: 2023-02-18
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    toc_float: true
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

A data structure is a unique way of organizing data so that it can be stored, processed, and retrieved effectively. In this lesson we will review some of the different data structures in R, including the following:

- Vectors
- Matrices
- Arrays

In the [Data Structure - Part 2 lesson](https://r4beginners.com/lessons/lesson5-data-structure2/){target="_blank"}, we will review: 

- Data frames
- Tibbles

Depending on what you are using R for, you will probably use a specific type of data structure most frequently. For example, in my research, I use data frames and tibbles all the time and rarely use matrices or arrays. However, it is still useful to be aware of and understand the different types of data structures.

## Vectors

In R, a vector is the simplest type of data structure. It is a sequence of data elements of the same basic type. 

In the example below, we have three people (who happen to be myself and my two brothers) - Josh, Jenny, and Brandon. Here, we are creating three separate vectors:

- `names` that contains *character strings* of our three names.
- `age` that contains *numeric values* representing our respective ages (at least at the time of writing this, but I will happily remain 30 forever).
- `blue_eyes` that contains *logical values* representing whether we have blue eyes or not.

```{r}
names <- c("Josh", "Jenny", "Brandon")
names

age <- c(31, 30, 27)
age

blue_eyes <- c(TRUE, FALSE, FALSE)
blue_eyes
```

Once you run the above R chunk, you can click on the Environment tab and see how the data is stored. It even shows the data types (num, logi, chr).

In this example, it's important to notice that each vector only contains one type of data. We can also see the type of data stored with the `class()` function. 

*The lines of code below are not being assigned (or saved) to any variables, so the results will be returned in the console, but not saved to the Environment.*

```{r}
class(names)
class(age)
class(blue_eyes)
```

## Matrices

A matrix has 2 dimensions of data and contains only one type of data. Matrices look like a typical table. In my experience, matrices typically contain numeric values, but there can also be character matrices. 

- R has a built-in `matrix()` function.
- In the example below, we will create a matrix of the numbers 1 through 25 and fill it into a 5x5 matrix (5 rows and 5 columns).
- We will save our matrix to a variable called `my_matrix`.

```{r}
my_matrix <- matrix(data = 1:25, nrow = 5, ncol = 5)
my_matrix
```

- We can see that the data perfectly fill in our matrix. 
- By default, R has filled in our matrix by columns, so we can see that 1, 2, 3, 4, 5 are filled in the first column and then it moves on to the next column.

If you want to fill in your matrix by rows (instead of columns), you can set the `byrow` *argument* to equal TRUE like in the example below.

*Please note, we will review functions and arguments in more detail in a couple of lessons.*

```{r}
my_matrix_byrow <- matrix(data = 1:25, nrow = 5, ncol = 5, byrow = TRUE)

my_matrix_byrow
```

Here's an example of a matrix with character strings, specifically the colors of the rainbow.

- Note, that the character strings are in quotes and wrapped around a `c()` function, which simply combines the elements.

```{r}
rainbow_matrix <- matrix(data = c("red", "orange", "yellow", 
                                  "green", "blue", "purple"), nrow = 2, ncol = 3)
rainbow_matrix
```

You can access an item within your matrix by using `[]` where the first number represents which column and the second represents which row.

```{r}
my_matrix[2,4]

my_matrix_byrow[2,4]

rainbow_matrix[1,3]
```

Next, we'll test out what happens if we try to create a matrix that is **smaller** than our data:

```{r}
matrix(data = 1:25, nrow = 4, ncol = 4)
```

- Here, we can see that R did print our 4x4 matrix, but we are also given a useful warning notifying us that the length of our data (25) does not fit

Next, let's test out what happens if we try to create a matrix that is **larger** than our given data:

```{r}
matrix(data = 1:25, nrow = 6, ncol = 6)
```

- Again, we see that R does return our matrix, but notice what happens after it reaches our max value of 25... it starts over! R will recycle the values, so after 25, it will go back to 1. 
- Note that we also get the same warning as we saw previously because our data does not fit perfectly.

Matrices are often used for data transformation. So as a final example of matrices, let's see how we can easily transform our `my_matrix` matrix to a new matrix that is multiplied by 2.

- This is as simple as multiplying our original matrix by 2!

```{r}
# original matrix
my_matrix

# matrix multiplied by 2
my_matrix*2
```

## Arrays

An array has 1 or more dimensions of data, but only contains a single data type. 

- R has a built-in `array()` function.
- Below, we will create a simple vector of a sequence 1 through 18 and.
- Then we will call the array function to see what happens.

*Even though we are calling an `array()` we can see that this type of an array is simply a vector.*

- Now we know that a 1-dimensional array is a vector.

```{r}
vector <- 1:25

array(vector)
```

- Let's delve a bit further!
- The `array()` function has an argument called "dim", which is where we can set the dimensions.
- Here are a few examples of multi-dimensional arrays:

```{r}
array(vector, dim = c(5,5))
```

- This should look very familiar! 
- This example is meant to demonstrate that 2 dimensional arrays are matrices.

However, arrays can also contain more than 2 dimensions...let's see what happens if we add another dimension.

```{r}
array(vector, dim = c(5,5,2))
```

- The first number in our dim argument tells us how many rows will be in our array 
- The second number specifies the number of columns
- The third number specifies the number of dimensions

Here's one final example, but I encourage you to play around with the `array()` and test out different numbers and dimensions

```{r}
array(vector, dim = c(2,3,4))
```

*For more examples check out [this website](https://www.w3schools.com/r/r_vectors.asp).*

## Summary 

In this lesson we introduced three types of data structures: vectors, matrices, and arrays. 

- A vector is the simplest type of data structure, where all the elements are the same data type and only has 1 dimension.

- A matrix also only contains elements of the same data type, but it has 2 dimensions.

- An array also only contains elements of the same data type, but an array can be as simple as a vector with 1 dimension or it can be more complex with several dimensions. 

## Exercises

1. Create a vector with three names: jennifer, taylor, and miley. Save this vector as `artists`. Print your results.

```{r}

```

What is the class of `artists`?

```{r}

```

2. Create a vector with the three artists heights in inches (Jennifer = 64.57, Taylor = 70.87, Miley = 64.96). Save the vector as `artists_heights`. Make sure to save the heights in the same order as you saved the names.

```{r}

```

What is the class of `artists_heights`?

```{r}

```

3. Create a matrix containing the values 1:10 with 2 rows and 5 columns.

```{r}

```

3. Create a matrix containing the values 1:10 with 5 rows and 2 columns.

```{r}

```

4. Create a matrix containing the values 1:10 with 5 rows and 2 columns, but this time fill in the values by rows (instead of the default which is by column).

```{r}

```

5. *Challenge Exercise* Create a matrix containing the first 10 letters in the alphabet with 5 rows and 2 columns and fill in the values by rows.

Hint: if you get stuck, try to use Google to learn how to print letters of the alphabet in R.

```{r}

```

## THE END `r emo::ji("tada")` 