---
title: "2. Variable Types"
description: |
   In this lesson we will introduce different types of variables.
# author:
#   - name: Jenny Sloane
# date: 2022-10-08
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

In this lesson we will introduce different types of variables, including character, numeric, integer, logical. 

## Types of Variables

### Character

- A character is a type of *class* that contains a string of characters. For example, a name is a string of characters (or letters).

- In the example below, I've saved my name "Jenny" to a variable I've called "my_char".

```{r}
my_char <- "Jenny"
```

- To check what type of variable you have, you can use the `class()` function
- All we have to do is insert our variable (my_char) within the parentheses of our function

*Note, we will review functions in a following lesson, but for now just run the below line of code and see what gets returned*

```{r}
class(my_char)
```

- As you can see, R returned the word "character" as we may have expected

- We can also use the `is.character()` function to check if our variable is a character. This function will return either TRUE or FALSE

```{r}
is.character(my_char)
```

### Numeric

- Numeric is a type of *class* that contains whole numbers *or* numbers with decimals
- Let's test out a couple of examples 
- We will first assign a number to a new variable (my_num1 or my_num2)
- And then we'll ask R to print (or return) the value for us

```{r}
my_num1 <- 22 # my favorite number!
my_num1 

my_num2 <- 22.22
my_num2
```

- Let's check the classes of the variables we've just created

```{r}
class(my_num1)
class(my_num2)
```

- We can see that both of our variables are numeric

### Integer

- An integer is a type of *class* that contains *only* whole numbers 
- Of course, this is similar to the numeric class, but without decimals. So how does R know if a whole number is numeric or an integer?
- By default, R will assign any whole number to the numeric class. We observed this above when we saw the class of the number 22 was numeric
- However, as we'll see below, you can convert a numeric class to an integer using the `as.integer()` function

```{r}
my_int1 <- as.integer(22)
class(my_int1)
```

- But, what happens if we try to convert a number with a decimal to an integer?

```{r}
my_int2 <- as.integer(22.22)
class(my_int2)
```

- This works! It has successfully converted 22.22 to an integer
- However, it is important to note that 22.22 was forced to become a whole number
- So, if we look at my_int2, it is now the whole number 22 rather than the decimal 22.22

```{r}
my_int2
```

### Logical

- Logical is a special type of *class* that contains only two possible values: TRUE or FALSE
- Please note that the logical values must be written in all caps with no quotes

```{r}
log1 <- TRUE
log2 <- FALSE
```

- Once again, we'll check the classes of the variables we've just created

```{r}
class(log1)
class(log2)
```

- What happens if we try the following?

```{r}
log3 <- "TRUE"
class(log3)
```

- Because TRUE is in "" here, by default R assigns this to the character class

```{r eval=FALSE}
log4 <- true
```

- This example doesn't even run! R gives us an error here saying "object 'true' not found"

## Summary

- Character: list of character strings 
- Numeric: whole numbers OR numbers with decimals
- Integer: Only whole numbers
- Logical: TRUE or FALSE

## THE END